---
title: "Lemur Disparity"
author: "April Lamb"
date: "5/13/2021"
output: html_document
---

Set working directory. Note that this will be different for each user.
```{r setup}
setwd('C:/Users/april/Desktop/Lemur')
```

Load libraries.
```{r load}
# Load libraries
pkgs <- c("phytools", "ape", "geiger", "dplyr", "devtools", "ggplot2", "coda", "diversitree", "nlme", "MCMCglmm", "PBSmapping", "siar", "spatstat", "sp", "rgdal", "rgeos")

vapply(pkgs, library, logical(1), character.only = TRUE, logical.return = TRUE, quietly = TRUE)
```

Load data
```{r data}
body <- read.csv("Lemurbody.csv") # Body size, traits, and PC axes 1 & 2
```

Draw convex hulls
```{r convex hull}

# Split based on Status
spx <- split(body$x,body$status)
spy <- split(body$x,body$status)


#now draw convex hulls around your data, subset by extinct, extant, and everything together
extant<-convexhull(spx[[1]],spy[[1]])
extant <- c(extant, extant[1])
all<-convexhull(fruit$x,fruit$y)
all <- c(all, all[1])
extinct<-convexhull(spx[[2]], spy[[2]] )
extinct <- c(extinct, extinct[1])


#now plot them so you can see
