---
title: "Lemur_Tooth_Shape"
author: "April Lamb"
date: "February 9 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### **Set up workspace** 

Set working directory. Note that this will be different for each user. 
```{r setup} 
setwd('C:/Users/april/Desktop/Lemur')
```
 
Load libraries and data.
```{r load, message=FALSE}
library(geiger)
library(devtools)
library(ggplot2)
devtools::install_github("MomX/Momocs", build_vignettes=TRUE)
library(Momocs)
library(dplyr)
library(knitr)
library(gridExtra)

load('/Users/april/Desktop/Lemur/Code/new_landmarked_shapes.RData') # shapefile data

data <- read.csv('/Users/april/Desktop/Lemur/Code/Lemur_KeyII.csv', header=T) # trait data
```

Source functions needed for analysis
```{r source}
source("/Users/april/Desktop/Lemur/Code/Lemur_functions.R")
```

#### **Aligning shapefiles**

```{r align, warning=FALSE}
# Combine your series of landmarks into a single landmarked series of objects
shapes_ldk <- c(shapes1,shapes2,shapes3,shapes4) 

# Use this series of functions to align all of your objects based on where the landmarks are. You might want to change this based on what it looks like.
aligned <- coo_bookstein(shapes_ldk,1,2)
aligned <- coo_slide(aligned, ldk=1)
aligned <- coo_rotate(aligned, (pi/2))

# This is the elliptical fourier analysis. Do this to check the alignment and make sure that everything works
Axis1F <- efourier(aligned, nb.h=30, norm=F)


# PCA on Fourier descriptors
Axis1D <- PCA(Axis1F)
plot(Axis1D) 
```

* At some point you'll want to append this information with your key to the teeth and species. it doesn't matter at which point. Probably you want to do this after you've landmarked everything and save it as a separate datafile.
* In additon, every momocs shape has a 'fac' component. When you create plots after your fourier analysis, you can color them by the'fac', or in this case your species or tooth type. You first read in your landmarked shapes and then you read in your excel file and then you specify the facs by the excel sheet values (see script below)

#### Elliptical Fourier Analysis 

```{r analysis, warning=FALSE} 

# First, lets make a tooth factor
tooth <- data$Tooth
tooth <- factor(tooth)
aligned$fac$col2 <- tooth

# Now make a species factor
species <- data$Species_name
species <- factor(species)
aligned$fac$col1 <- species

# Take a look at your aligned dataframe:
head(aligned$fac) # To check the first 6 rows
tail(aligned$fac) # To check the last 6 rows
aligned$fac[25:30,] # To check specific rows

# Pull out trait data and make each a factor. Then move these into the fac dataframe in appropriate columns.

# Diet
diet <- data$Diet 
diet <- factor(diet)
aligned$fac$col3 <- diet 
head(aligned$fac) 

# Habitat
habitat <- data$Habitat 
habitat <- factor(habitat)
aligned$fac$col4 <- habitat
head(aligned$fac)

# Status 1
extinctextant <- data$Extinct_Extant
extinctextant <- factor(extinctextant)
aligned$fac$col5 <- extinctextant
head(aligned$fac)

# IUCN
iucn <- data$IUCN
iucn <- factor(iucn)
aligned$fac$col6 <- iucn
head(aligned$fac)

# Remove all of the outgroups i.e. everything that DOESN'T live in Madagascar and other unwanted species
rm_out <- Momocs::filter(aligned, !aligned$fac$col1=="Otolemur_monteiri")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Daubentonia_madagascariensis")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Archaeolemur")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Galago_senegalensis")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Hadropithecus")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Lepilemur")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Loris_tardigradus")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Nycticebus")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Nycticebus_coucang")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Nycticebus_javanicus")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Nycticebus_menagensis")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Nycticebus_pygmaeus")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Megaladapis")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Mesopropithecus")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Microcebus")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Otolemur")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Otolemur_crassicaudatus")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Pachylemur_insignis")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Palaeopropithecus")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Palaeopropithecus_kelyus")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Propithecus")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Perodicticus_potto")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Perodicticus_edwardsi")
rm_out <- Momocs::filter(rm_out, !rm_out$fac$col1=="Prolemur_simus")

# Divide data by tooth type to see if there are any broad differences. Note that for the purposes of this analysis we won't be looking at the tooth comb.
molar <- Momocs::filter(rm_out, !rm_out$fac$col2 =="TC") # All molars, excluding the tooth combs
LR_molar <- Momocs::filter(rm_out, rm_out$fac$col2 %in% c("LR")) # lower right molar
LL_molar <- Momocs::filter(rm_out, rm_out$fac$col2 %in% c("LL")) # lower left molar
UR_molar <- Momocs::filter(rm_out, rm_out$fac$col2 %in% c("UR")) # upper right molar
UL_molar <- Momocs::filter(rm_out, rm_out$fac$col2 %in% c("UL")) # upper left molar

# Stack everything in an object and color it by factor 
stack(molar, fac="col1")
stack(UR_molar, fac="col1")
stack(LR_molar, fac="col1")
stack(UL_molar, fac="col1")
stack(LL_molar, fac="col1")

# Now perform the elliptical fourier analysis
UR <- efourier(UR_molar, nb.h=30, norm=F)
LR <- efourier(LR_molar, nb.h=30, norm=F) 
LL <- efourier(LL_molar, nb.h=30, norm=F)
UL <- efourier(UL_molar, nb.h=30, norm=F)
mol <- efourier(molar, nb.h=30, norm=F)

```

#### **Principal Component Analysis (PCA)** 

```{r PCA, warning=FALSE, results="hide"}
# Run PCA by tooth type 
LRpca <- PCA(LR)
URpca <- PCA(UR)
LLpca <- PCA(LL)
ULpca <- PCA(UL)
molpca <- PCA(mol)

# Get loadings
# Lower right molar
LRx <- LRpca$x
LRx <- as.data.frame(LRpca$x)
LRx <- tibble::rownames_to_column(LRx, "Photo_ID")

# Lower left molar
LLx <- LLpca$x
LLx <- as.data.frame(LLpca$x)
LLx <- tibble::rownames_to_column(LLx, "Photo_ID")

# Upper right molar
URx <- URpca$x
URx <- as.data.frame(URpca$x)
URx <- tibble::rownames_to_column(URx, "Photo_ID")
 
# Upper left molar
ULx <- ULpca$x
ULx <- as.data.frame(ULpca$x)
ULx <- tibble::rownames_to_column(ULx, "Photo_ID")

# All molars
molx <- molpca$x
molx <- as.data.frame(molpca$x)
molx <- tibble::rownames_to_column(molx, "Photo_ID")

# For some reason there are discrepencies in the photo names between files (???) so we'll quickly fix those here. Otherwise the final PCA files will contain NA's bc they won't join correctly with the trait file.
# LRx
LRx1 <- as.data.frame(sapply(LRx[,1], gsub, pattern="-", replacement=".")) # Pull out names col & fix errors
as.list(LRx1)
names(LRx1)[1]<-paste("Photo_ID") # Rename column
LRsub <- subset(LRx, select = -c(Photo_ID)) # Subset original df to exclude wrong names col
LRx <- cbind(LRx1, LRsub) # Merge correct names column w/ original df

# LLx
LLx1 <- as.data.frame(sapply(LLx[,1], gsub, pattern="-", replacement=".")) # Pull out names col & fix errors
as.list(LLx1)
names(LLx1)[1]<-paste("Photo_ID") # Rename column
LLsub <- subset(LLx, select = -c(Photo_ID)) # Subset original df to exclude wrong names col
LLx <- cbind(LLx1, LLsub) # Merge correct names column w/ original df

#URx
URx1 <- as.data.frame(sapply(URx[,1], gsub, pattern="-", replacement=".")) # Pull out names col & fix errors
as.list(URx1)
names(URx1)[1]<-paste("Photo_ID") # Rename column
URsub <- subset(URx, select = -c(Photo_ID)) # Subset original df to exclude wrong names col
URx <- cbind(URx1, URsub) # Merge correct names column w/ original df

#ULx
ULx1 <- as.data.frame(sapply(ULx[,1], gsub, pattern="-", replacement=".")) # Pull out names col & fix errors
as.list(ULx1)
names(ULx1)[1]<-paste("Photo_ID") # Rename column
ULsub <- subset(ULx, select = -c(Photo_ID)) # Subset original df to exclude wrong names col
ULx <- cbind(ULx1, ULsub) # Merge correct names column w/ original df

# Now add the other factors
LRx <- left_join(LRx, data)
LLx <- left_join(LLx, data)
URx <- left_join(URx, data)
ULx <- left_join(ULx, data)
molx <- left_join(molx, data)

# Export PCA results 
write.csv(LRx, "Lower_right_molar_PCA.csv", quote=F)
write.csv(LLx, "Lower_left_molar_PCA.csv", quote=F)
write.csv(URx, "Upper_right_molar_PCA.csv", quote=F)
write.csv(ULx, "Upper_left_molar_PCA.csv", quote=F)
write.csv(molx, "All_molars_PCA.csv", quote=F)

# Plot results colored by factor
# Diet
# pdf("DietxTeeth.pdf")
plot(molpca, fac=molpca$fac$col3, ellipses=T, labelgroups=F, title = "molpca x diet") 
plot(LRpca, fac=LRpca$fac$col3, ellipses=T, labelgroups=F, title = "LRpca x diet") 
plot(LLpca, fac=LLpca$fac$col3, ellipses=T, labelgroups=F, title = "LLpca x diet") 
plot(URpca, fac=URpca$fac$col3, ellipses=T, labelgroups=F, title = "URpca x diet") 
plot(ULpca, fac=ULpca$fac$col3, ellipses=T, labelgroups=F, title = "ULpca x diet") 
#dev.off()


# Habitat
# pdf("HabitatxTeeth.pdf")
plot(molpca, fac=molpca$fac$col4, ellipses=T, labelgroups=F, title = "molpca x habitat") 
plot(LRpca, fac=LRpca$fac$col4, ellipses=T, labelgroups=F, title = "LRpca x habitat") 
plot(LLpca, fac=LLpca$fac$col4, ellipses=T, labelgroups=F, title = "LLpca x habitat") 
plot(URpca, fac=URpca$fac$col4, ellipses=T, labelgroups=F, title = "URpca x habitat") 
plot(ULpca, fac=ULpca$fac$col4, ellipses=T, labelgroups=F, title = "ULpca x habitat") 
#dev.off()


# Status (Extinct v. Extant)
# pdf("StatusxTeeth.pdf")
plot(molpca, fac=molpca$fac$col5, ellipses=T, labelgroups=F, title = "molpca x status") 
plot(LRpca, fac=LRpca$fac$col5, ellipses=T, labelgroups=F, title = "LRpca x status") 
plot(LLpca, fac=LLpca$fac$col5, ellipses=T, labelgroups=F, title = "LLpca x status") 
plot(URpca, fac=URpca$fac$col5, ellipses=T, labelgroups=F, title = "URpca x status") 
plot(ULpca, fac=ULpca$fac$col5, ellipses=T, labelgroups=F, title = "ULpca x status") 
# dev.off()

# IUCN Status
#pdf("IUCNxTeeth.pdf")
plot(molpca, fac=molpca$fac$col6, ellipses=T, labelgroups=F, title = "molpca x IUCN") 
plot(LRpca, fac=LRpca$fac$col6, ellipses=T, labelgroups=F, title = "LRpca x IUCN") 
plot(LLpca, fac=LLpca$fac$col6, ellipses=T, labelgroups=F, title = "LLpca x IUCN") 
plot(URpca, fac=URpca$fac$col6, ellipses=T, labelgroups=F, title = "URpca x IUCN") 
plot(ULpca, fac=ULpca$fac$col6, ellipses=T, labelgroups=F, title = "ULpca x IUCN") 
#dev.off()


# Get PCA means for each tooth type across species
# All molars
targets <- unique(molx$Species_name) # Identify targets
PC1 <- getMean(targets, molx, axis_column = 2) # get means
PC2 <- getMean(targets, molx, axis_column = 3)
PC3 <- getMean(targets, molx, axis_column = 4)
PC4 <- getMean(targets, molx, axis_column = 5)
PC5 <- getMean(targets, molx, axis_column = 6)
AllMolars <- cbind(PC1,PC2,PC3,PC4,PC5) # combine
write.csv(AllMolars, "allMolars.csv") # export

#Lower right Molar
targets<-unique(LRx$Species_name) # Identify targets
PC1 <- getMean(targets, LRx, axis_column = 2) # get means
PC2 <- getMean(targets, LRx, axis_column = 3)
PC3 <- getMean(targets, LRx, axis_column = 4)
PC4 <- getMean(targets, LRx, axis_column = 5)
PC5 <- getMean(targets, LRx, axis_column = 6)
lowerRightMolars <- cbind(PC1,PC2,PC3,PC4,PC5) # combine
write.csv(lowerRightMolars, "lowerRight.csv") # export

# Lower left molar
targets <- unique(LLx$Species_name) # Identify targets
PC1 <- getMean(targets, LLx, axis_column = ) # get means
PC2 <- getMean(targets, LLx, axis_column = 3)
PC3 <- getMean(targets, LLx, axis_column = 4)
PC4 <- getMean(targets, LLx, axis_column = 5)
PC5 <- getMean(targets, LLx, axis_column = 6)
lowerLeftMolars <- cbind(PC1,PC2,PC3,PC4,PC5) # combine
write.csv(lowerLeftMolars, "lowerLeft.csv") # export

# Upper right molar
targets <- unique(URx$Species_name) # Identify targets
PC1 <- getMean(targets, URx, axis_column = 2) # get means
PC2 <- getMean(targets, URx, axis_column = 3)
PC3 <- getMean(targets, URx, axis_column = 4)
PC4 <- getMean(targets, URx, axis_column = 5)
PC5 <- getMean(targets, URx, axis_column = 6)
upperRightMolars <- cbind(PC1,PC2,PC3,PC4,PC5) # combine
write.csv(upperRightMolars, "upperRight.csv") # export

# Upper left molar
targets <- unique(ULx$Species_name) # Identify targets
PC1 <- getMean(targets, ULx, axis_column = 2) # get means
PC2 <- getMean(targets, ULx, axis_column = 3)
PC3 <- getMean(targets, ULx, axis_column = 4)
PC4 <- getMean(targets, ULx, axis_column = 5)
PC5 <- getMean(targets, ULx, axis_column = 6)
upperLeftMolars <- cbind(PC1,PC2,PC3,PC4,PC5) # combine
write.csv(upperLeftMolars, "upperLeft.csv") # export
```






